Sparse Matrix Vector Multiplication with Multiple vectors and Multiple Matrices on the MIC Architecture

### 2013-10-16 ieee_paper.tex
Modified source code with omnigraffle. Why doesn't figure span both columns?

The RBF differentiation matrix, $D_N$, is derived by applying the desired analytic derivative operator $L$ to the RBF interpolant $s(\xvec)$ above and evaluating it at the point locations. For very large problems, this is a computationally expensive since the matrix in (\ref{eq:rbf}) is full and inversion requires O$(N^3)$ operations. To alleviate the cost of this global approach (i.e. using every node in the domain to calculate the derivative at a given node $\xvec_i$), RBF-generated finite differences (RBF-FD) have been derived \cite{TAI1,TAI2,SDY02,WrFo06,FoL11,FLBWSC12}. RBF-FD use only a local set of the $n_z-1$ nearest neighbors to the point $\xvec_i$ to approximate the derivative. In other words, $Lf(\xvec_i)=\sum_{j=1}^{n_z}a_jf(\xvec_j)$. The differentiation weights, $a_j$, are calculated by enforcing that this linear combination should be exact for RBFs, $\{\phi(\|\xvec-\xvec_{j}\|)\}_{j=1}^n_z$, centered at each of the node locations $\{\xvec_j\}_{j=1}^{n_z}$ (classical finite differences (FD) would enforce that it be exact for polynomials instead). Similar to FD, as the stencil size $n_z$ increases so does the order of the method.

@@

**** PUT ELSEWHERE OR ELIMINATE? ****

When solving a one dimensional system of PDEs, one might require an
$x$ derivative of multiple functions. For example, the Euler equations
require the $x$ derivative of the three components of velocity and
pressure. In this case, there are $n_v=4$ vectors $x^k$,
$k=0,\cdots,3$.

Thus for each vector element $y_i$, we compute $y_i = \sum_j A_{ij}
x_j$. If $A_i$ is row $i$ of $A$, $y_i$ is simply the dot product $A_i
x$. The next level of generality is to consider $n_v$ vectors $x^k$,
$k=0,\cdots,n_v-1$. Whatever the spmv implementation, one achieves
improved performance if the matrix formed from the columns $x^k$ are
stored in row major order. Thus, $x^0_0,x^1_0,\cdots,x^{n_v-1}_0$, are
stored in consecutive memory location. The random access of the
elements of $x$ is thus reduced. Maximum efficiency is achieved when
$n_v=16$ floats or $8$ doubles, given that cache lines take 64
bytes. We will benchmark this case, labeled $Svn$, where $n$ refers to
the number of vectors (Iv4 uses four vectors). The $S$ refers to singe
precision. A double precision run is labelled $Dvn$.

Alternatively, when solving a PDE, one might require derivatives of a
given scalar function with respect to coordinate directions $x$, $y$,
$z$. Second order operators of often required, such as a second
derivative with respect to $x$ or a Laplacian operator. In the RBFFD
formulation, on can compute different derivatives using the same
stencil, but with different weights. In other words, the adjacency
matrix that corresponds to $A$ remains constant, but the matrix
elements of $A$ change with the particular derivative.  In this case,
label with a superscript $l$ the particular matrix $A^k$. Since the
adjacency matrix is assumed invariant, there is only need for a single
matrix \ttt{C{ij}}. In Ellpack format, each row is of constant size
(the number of nonzeros per row of $A$. $C_{ij}$ is the column number
that locates the $j^{th}$ nonzero in row $i$ of $A$.

\todo{Must rewrite the above to make it more focused on the
  application. Remove the architectural/implementation details such as
  using ELLPACK. Let's focus on on why this computation is important
  and how does the 16 multiplication appears in the equations.}

@@

\begin{figure}[h]
\begin{center}
\lstset{language=c++,basicstyle=\scriptsize,numbers=left,numbersep=5pt}
\begin{lstlisting}{float=h, Caption}
const __m512i offsets = _mm512_set4_epi32(3,2,1,0);
const __m512i four = _mm512_set4_epi32(4,4,4,4); 
int int_mask = 0x1111;
__mmask16 mask = _mm512_int2mask(int_mask);

#pragma omp parallel for
for (int r=0; r < nb_rows; r++) {
  __m512 accu = _mm512_setzero_ps();

  for (int n=0; n < nz; n+=4) {
    float* a = &dom.col_id_t[0]+n+nz*r;
    __m512i vect_i = _mm512_setzero_ps();
    vect_i = _mm512_mask_loadunpacklo_ps(vect_i, mask, a);
    vect_i = _mm512_mask_loadunpackhi_ps(vect_i, mask, a);
    vect_i = _mm512_swizzle_ps(vect_i, _MM_SWIZ_REG_AAAA);
    vect_i = _mm512_fmadd_epi32(vect_i, four, offsets);

    all_vect = _mm512_i32gather_ps(vect_i, dom.vec_vt, scale);

    mat_ent = _mm512_load_ps(dom.data_t + nb_mat*(n + r*nz));

    //perform first tensor multiplication
    correct_vect = permute(all_vect, _MM_PERM_AAAA);
    correct_mat = _mm512_swizzle_ps(mat_ent, _MM_SWIZ_REG_AAAA);
    accu = _mm512_fmadd_ps(correct_vect, correct_mat, accu);

    //three more times with BBBB, CCCC, and DDDD
  }
  _mm512_storenrngo_ps(dom.result_vt+nb_mat*nb_vec*r, accu);  
\end{lstlisting}
\end{center}
\end{figure}

@@

\begin{figure*}
  \centering

  \subfigure[Performance of $y=Ax$ on the MIC. Squares: base 1/1 case,
    solid circles: 4/4 case implemented in C++, solid triangles: 4/4
    case implemented with MIC vector instructions. $-O3$ compilation
    options. Each symbol has two cases: with (red) and without blue) 
    with and without Reverse Cuthill-McKee. The grid is $96^3$.}
    {\includegraphics[width=.49\linewidth]{figures/mic_performance_nb_threads.pdf}\label{}}
%
  \subfigure[performance on the host on a sparse matrix with $64^3$ rows. Manual
    vectoization brings significant improvement there as
    well.]{\includegraphics[width=.49\linewidth]{figures/64x64x64.pdf}\label{}}
  \caption{Vectorization matters in practice. Differences in BW
    translate in different performance.\todo{this two plots should ``correlate'' }}
\end{figure*}

>>>>

The RBF differentiation matrix, $D_N$, is derived by applying the desired analytic derivative operator $L$ to the RBF interpolant $s(\xvec)$ above and evaluating it at the point locations. For very large problems, this is a computationally expensive since the matrix in (\ref{eq:rbf}) is full and inversion requires O$(N^3)$ operations. To alleviate the cost of this global approach (i.e. using every node in the domain to calculate the derivative at a given node $\xvec_i$), RBF-generated finite differences (RBF-FD) have been derived \cite{TAI1,TAI2,SDY02,WrFo06,FoL11,FLBWSC12}. RBF-FD use only a local set of the $n_z-1$ nearest neighbors to the point $\xvec_i$ to approximate the derivative. In other words, $Lf(\xvec_i)=\sum_{j=1}^{n_z}a_jf(\xvec_j)$. The differentiation weights, $a_j$, are calculated by enforcing that this linear combination should be exact for RBFs, $\{\phi(\|\xvec-\xvec_{j}\|)\}_{j=1}^{n_z}$, centered at each of the node locations $\{\xvec_j\}_{j=1}^{n_z}$ (classical finite differences (FD) would enforce that it be exact for polynomials instead). Similar to FD, as the stencil size $n_z$ increases so does the order of the method.

@@

%**** PUT ELSEWHERE OR ELIMINATE? ****
%
%When solving a one dimensional system of PDEs, one might require an
%$x$ derivative of multiple functions. For example, the Euler equations
%require the $x$ derivative of the three components of velocity and
%pressure. In this case, there are $n_v=4$ vectors $x^k$,
%$k=0,\cdots,3$.
%
%Thus for each vector element $y_i$, we compute $y_i = \sum_j A_{ij}
%x_j$. If $A_i$ is row $i$ of $A$, $y_i$ is simply the dot product $A_i
%x$. The next level of generality is to consider $n_v$ vectors $x^k$,
%$k=0,\cdots,n_v-1$. Whatever the spmv implementation, one achieves
%improved performance if the matrix formed from the columns $x^k$ are
%stored in row major order. Thus, $x^0_0,x^1_0,\cdots,x^{n_v-1}_0$, are
%stored in consecutive memory location. The random access of the
%elements of $x$ is thus reduced. Maximum efficiency is achieved when
%$n_v=16$ floats or $8$ doubles, given that cache lines take 64
%bytes. We will benchmark this case, labeled $Svn$, where $n$ refers to
%the number of vectors (Iv4 uses four vectors). The $S$ refers to singe
%precision. A double precision run is labelled $Dvn$.
%
%Alternatively, when solving a PDE, one might require derivatives of a
%given scalar function with respect to coordinate directions $x$, $y$,
%$z$. Second order operators of often required, such as a second
%derivative with respect to $x$ or a Laplacian operator. In the RBFFD
%formulation, on can compute different derivatives using the same
%stencil, but with different weights. In other words, the adjacency
%matrix that corresponds to $A$ remains constant, but the matrix
%elements of $A$ change with the particular derivative.  In this case,
%label with a superscript $l$ the particular matrix $A^k$. Since the
%adjacency matrix is assumed invariant, there is only need for a single
%matrix \ttt{C{ij}}. In Ellpack format, each row is of constant size
%(the number of nonzeros per row of $A$. $C_{ij}$ is the column number
%that locates the $j^{th}$ nonzero in row $i$ of $A$.
%
%
%\todo{Must rewrite the above to make it more focused on the
% application. Remove the architectural/implementation details such as
%  using ELLPACK. Let's focus on on why this computation is important
%  and how does the 16 multiplication appears in the equations.}

@@

\newfig{figures/source_code.pdf}{6in}{Source code}{fig:source}

%\begin{figure}[h]
%\begin{center}
%\lstset{language=c++,basicstyle=\scriptsize,numbers=left,numbersep=5pt}
%\begin{lstlisting}{float=h, Caption}
%const __m512i offsets = _mm512_set4_epi32(3,2,1,0);
%const __m512i four = _mm512_set4_epi32(4,4,4,4); 
%int int_mask = 0x1111;
%__mmask16 mask = _mm512_int2mask(int_mask);
%
%#pragma omp parallel for
%for (int r=0; r < nb_rows; r++) {
%  __m512 accu = _mm512_setzero_ps();
%
%  for (int n=0; n < nz; n+=4) {
%    float* a = &dom.col_id_t[0]+n+nz*r;
%    __m512i vect_i = _mm512_setzero_ps();
%    vect_i = _mm512_mask_loadunpacklo_ps(vect_i, mask, a);
%    vect_i = _mm512_mask_loadunpackhi_ps(vect_i, mask, a);
%    vect_i = _mm512_swizzle_ps(vect_i, _MM_SWIZ_REG_AAAA);
%    vect_i = _mm512_fmadd_epi32(vect_i, four, offsets);
%
%    all_vect = _mm512_i32gather_ps(vect_i, dom.vec_vt, scale);
%
%    mat_ent = _mm512_load_ps(dom.data_t + nb_mat*(n + r*nz));
%
%    //perform first tensor multiplication
%    correct_vect = permute(all_vect, _MM_PERM_AAAA);
%    correct_mat = _mm512_swizzle_ps(mat_ent, _MM_SWIZ_REG_AAAA);
%    accu = _mm512_fmadd_ps(correct_vect, correct_mat, accu);
%
%    //three more times with BBBB, CCCC, and DDDD
%  }
%  _mm512_storenrngo_ps(dom.result_vt+nb_mat*nb_vec*r, accu);  
%\end{lstlisting}
%\end{center}
%\end{figure}
%

@@

\begin{figure*}
  \centering
  \subfigure[Performance of $y=Ax$ on the MIC. Squares: base 1/1 case],
    solid circles: 4/4 case implemented in C++, solid triangles: 4/4
    case implemented with MIC vector instructions. $-O3$ compilation
    options. Each symbol has two cases: with (red) and without blue) 
    with and without Reverse Cuthill-McKee. The grid is $96^3$.}
    {\includegraphics[width=.49\linewidth]{figures/mic_performance_nb_threads.pdf}\label{fig:perf_mic}}
%
  \subfigure[performance on the host on a sparse matrix with $64^3$ rows. Manual
    vectorization brings significant improvement there as
    well.]{\includegraphics[width=.49\linewidth]{figures/64x64x64.pdf}\label{fig:perf_host}}
  \caption{Vectorization matters in practice. Differences in BW
    translate in different performance.\todo{this two plots should ``correlate'' }}
\end{figure*}
